---
title: "ESS 330 Lab 4: Statistical Analysis in R"
author: "Erica Mortland"
format: 
  html: 
    self-contained: true
execute:
  echo: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries
```{r}
library(tidyverse)
library(ggpubr)
library(lterdatasampler)
library(car)
library(visdat)
remotes::install_github("lter/lterdatasampler")
```

# Load Dataset
```{r}
data("and_vertebrates")
glimpse(and_vertebrates)
```

## 1. Chi-Square Test: Forest Type & Channel Unit (Coastal Giant Salamander)
```{r}
salamander_data <- and_vertebrates %>% 
  filter(species == "Coastal giant salamander") %>% 
  drop_na(unittype, section)

cont_table <- table(salamander_data$section, salamander_data$unittype)
chisq.test(cont_table)
```
Findings: The chi-square test shows a significant relationship between forest type and channel unit (X² = r chisq_test_result$statistic, p = r chisq_test_result$p.value), indicating these variables are aren't independent.

## 2. T-test: Biomass Differences in Coastal Giant Salamander
```{r}
salamander_data <- salamander_data %>% drop_na(weight_g)

t.test(log(salamander_data$weight_g) ~ salamander_data$section, var.equal = TRUE)
```
Findings: The t-test suggests a difference in biomass between forest types (t = r t_test_result$statistic, p = r t_test_result$p.value). The log-transformed biomass was higher in one forest type.

## 3. Correlation Test: Length vs. Weight (Cutthroat Trout)
```{r}
trout_data <- and_vertebrates %>% 
  filter(species == "Cutthroat trout") %>% 
  drop_na(length_1_mm, weight_g)

cor.test(log(trout_data$length_1_mm), log(trout_data$weight_g))
```
Findings: The Pearson correlation test shows a strong positive relationship between length and weight (r = r correlation_result$estimate, p = r correlation_result$p.value), indicating larger fish tend to weigh more.

# Part 2: Multivariate Statistics
## Load Fiddler Crab Dataset
```{r}
data("pie_crab")
glimpse(pie_crab)
```

## 4. ANOVA & Tukey’s HSD: Crab Size Across Sites
```{r}
pie_anova <- aov(size ~ site, data = pie_crab)
summary(pie_anova)
TukeyHSD(pie_anova)
```
Findings: ANOVA shows a  difference in crab size across sites (F = r summary(pie_anova)[[1]][["F value"]][1], p = r summary(pie_anova)[[1]][["Pr(>F)"]][1]). Tukey's test reveals which sites differ significantly.

## 5. Simple Linear Regression: Water Temp SD vs. Carapace Width
```{r}
pie_lm <- lm(size ~ water_temp_sd, data = pie_crab)
summary(pie_lm)
```

## 6. Multiple Linear Regression
```{r}
cor(pie_crab %>% select(latitude, air_temp_sd, water_temp_sd))
pie_mlm <- lm(size ~ latitude + air_temp_sd + water_temp_sd, data = pie_crab)
summary(pie_mlm)
```

```{r}

pie_crab <- pie_crab %>%
  distinct(site, .keep_all = TRUE) %>%
  arrange(latitude) %>%
  mutate(site = factor(site, levels = unique(site)))  # Set levels based on unique sites in order of latitude

# Plot the boxplot
ggplot(pie_crab, aes(x = site, y = size)) +
  geom_boxplot() +
  xlab("Site") +
  ylab("Carapace Width (size)") +
  ggtitle("Carapace Width Across Sites Ordered by Latitude") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


```{r}

pie_anova <- aov(size ~ site, data = pie_crab)
summary(pie_anova)

TukeyHSD(pie_anova)

```
```{r}
pie_lm <- lm(size ~ water_temp_sd, data = pie_crab)
summary(pie_lm)

ggplot(pie_crab, aes(x = water_temp_sd, y = size)) +
  geom_point() +
  geom_smooth(method = "lm", col = "blue") +
  xlab("Water Temperature Standard Deviation") +
  ylab("Carapace Width (size)") +
  ggtitle("Simple Linear Regression: Water Temperature SD vs. Carapace Width")
```

```{r}
cor(pie_crab %>% select(latitude, air_temp_sd, water_temp_sd))

pie_mlm <- lm(size ~ latitude + air_temp_sd + water_temp_sd, data = pie_crab)
summary(pie_mlm)
```



The regression confirms that latitude, air temperature variation, and water temperature variation significantly influence crab size.
