---
title: "ESS 330 Lab 4: Statistical Analysis in R"
author: "Your Name"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Load Libraries
```{r}
library(tidyverse)
library(ggpubr)
library(lterdatasampler)
library(car)
library(visdat)
remotes::install_github("lter/lterdatasampler")
```

# Load Dataset
```{r}
data("and_vertebrates")
glimpse(and_vertebrates)
```

## 1. Chi-Square Test: Forest Type & Channel Unit (Coastal Giant Salamander)
```{r}
salamander_data <- and_vertebrates %>% 
  filter(species == "Coastal giant salamander") %>% 
  drop_na(unittype, section)

cont_table <- table(salamander_data$section, salamander_data$unittype)
chisq.test(cont_table)
```

## 2. T-test: Biomass Differences in Coastal Giant Salamander
```{r}
salamander_data <- salamander_data %>% drop_na(weight_g)

t.test(log(salamander_data$weight_g) ~ salamander_data$section, var.equal = TRUE)
```

## 3. Correlation Test: Length vs. Weight (Cutthroat Trout)
```{r}
trout_data <- and_vertebrates %>% 
  filter(species == "Cutthroat trout") %>% 
  drop_na(length_1_mm, weight_g)

cor.test(log(trout_data$length_1_mm), log(trout_data$weight_g))
```

# Part 2: Multivariate Statistics
## Load Fiddler Crab Dataset
```{r}
data("pie_crab")
glimpse(pie_crab)
```

## 4. ANOVA & Tukey’s HSD: Crab Size Across Sites
```{r}
pie_anova <- aov(size ~ site, data = pie_crab)
summary(pie_anova)
TukeyHSD(pie_anova)
```

## 5. Simple Linear Regression: Water Temp SD vs. Carapace Width
```{r}
pie_lm <- lm(size ~ water_temp_sd, data = pie_crab)
summary(pie_lm)
```

## 6. Multiple Linear Regression
```{r}
cor(pie_crab %>% select(latitude, air_temp_sd, water_temp_sd))
pie_mlm <- lm(size ~ latitude + air_temp_sd + water_temp_sd, data = pie_crab)
summary(pie_mlm)
```